---
title: Analysis of historic data to identify areas in which resources should be prioritised
  in the event of a storm.
author: "OfficialBenWhite"
date: "20 July 2015"
output: 
html_document:
    keep_md: true
---

The aim of this assignment was to investigate the NOAA storm database and try to make inferences using historic data, about how resources should be allocated to lessen the damage of a environmental event in the future.

##Data Processing

Load packages
```{r, results='hide'}
library(ggplot2)
library(lubridate)
library(reshape2)
library(dplyr)
```

Next download the storm data from the NOAA Storm Database

Set the working directory, the file variable and the url.

```{r}
setwd("/Users/bw/RCourse/C5W3")
file <- "StormData.csv.bz2"
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
```

Checks if file is downloaded otherwise download it.

```{r}
if(!file.exists(file)){   
    download.file(url,file, method="curl")
}
```

Use read.csv() to read the file into memory. 

```{r, cache=TRUE}
stormData <- read.csv(file)
```

As per this link: 
https://www.ncdc.noaa.gov/stormevents/details.jsp
Filter out any event that happened before 1996 because not all events were accurately recorded before then.

```{r}
stormData$date<- mdy_hms(stormData$BGN_DATE)
stormData2 <- subset(stormData, stormData$date > mdy("12/31/1995"))
```

Combine the PROPDMG and CROPDMG columns with their multipliers.
Page 12 of https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf states that:" Alphabetical characters used to signify magnitude include “K” for thousands, “M” for millions, and “B” for billions. "
Looking at each value and remove any anomolies.

```{r}
table(stormData2$PROPDMGEXP)
```

Therefore subsetting out everything that is not K , M, B or blank in the exponential columns.

Subset on the good values.

```{r}
indx  <- stormData2$CROPDMGEXP %in% c("K","M","B"," ", "")
indx2 <- stormData2$PROPDMGEXP  %in% c("K","M","B"," ", "")
importantColumns <- c("date", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
df <- stormData2[(indx && indx2), importantColumns]
```


Translate the exponents into numbers.

```{r}
df$cropdmgexpn[df$CROPDMGEXP == "K"] <- 1000
df$cropdmgexpn[df$CROPDMGEXP == "M"] <- 1e+06
df$cropdmgexpn[df$CROPDMGEXP == "B"] <- 1e+09
df$propdmgexpn[df$PROPDMGEXP == "K"] <- 1000
df$propdmgexpn[df$PROPDMGEXP == "M"] <- 1e+06
df$propdmgexpn[df$PROPDMGEXP == "B"] <- 1e+09
```

Create new columns for crop damage and property damage with exponents included.

```{r}
df$cropdmg = df$cropdmgexpn * df$CROPDMG
df$propdmg = df$propdmgexpn * df$PROPDMG
```

Now group by EVTYPE and sum the data. Also add relative frequency and count.

```{r}
grp <- group_by(df, EVTYPE)
sumdf <- summarise(grp, fatalities = sum(FATALITIES, na.rm = TRUE), injuries = sum(INJURIES, na.rm = TRUE), cropdamage = sum(cropdmg, na.rm = TRUE), propdamage = sum(propdmg, na.rm = TRUE), count = n())

sumdf <- df %>%
    group_by(EVTYPE)  %>%
    summarise(fatalities = sum(FATALITIES, na.rm = TRUE), 
              injuries = sum(INJURIES, na.rm = TRUE), 
              cropdamage = sum(cropdmg, na.rm = TRUE), 
              propdamage = sum(propdmg, na.rm = TRUE),
              totaldamage = (sum(propdmg, na.rm = TRUE) + 
                                 sum(cropdmg, na.rm = TRUE)),
              count = n())  %>%
    mutate(rel.freq = paste0(round(100 * count/sum(count), 0), "%"))
```

##Results

The most fatal events:

```{r, results='asis'}
fatals <- head(arrange(sumdf, desc(fatalities)), n = 6 )
fatal <- fatals$EVTYPE[1]
knitr::kable(fatals)
```

As can be seen `r fatal` is the most fatal event type.

Now by injury

```{r, results='asis'}
knitr::kable(head(arrange(sumdf, desc(injuries)), n = 6 ) )
```

Now by total damage

```{r, results='asis'}
knitr::kable(head(arrange(sumdf, desc(totaldamage)), n = 6 ) )
```

Plot a chart of total damage with the fill as the the damage type.

```{r, fig.align='center'}
df2 <- head(arrange(sumdf, desc(totaldamage)), n = 10)
df3 <- df2[,c("EVTYPE", "propdamage", "cropdamage")]
df4 <- melt(df3, id.vars = "EVTYPE")
ggplot(df4, aes(x = reorder(EVTYPE,-value), y = value/1000000, fill = variable)) + 
    geom_bar(stat = "identity") +
    xlab("Event Type") +
    theme(axis.text.x = element_text(angle = 45, size=8, 
                                     hjust = 1, vjust = 1)) +
     ylab("Total Damage (millions of USD)") +
    ggtitle("Aggregated property and crop damage \nfor weather events from 1996-2011")
    
```

##Conclusion

As can be seen from the data, floods are catastrophic in terms of their lethalness and destructiveness. Perhaps its wise to allocate available resources to preventative measures such as building flood protection.